ScPy: A SuperCollider Extension for Performing Numerical Computation via Embedded Python

### 2016-08-29 doc/paper.tex
edit run 2

Nilson~\cite{nilson2007live} presents a review of the challenges of live coding practice. Live
coding presents an interesting contrast to contemporary music with many clear differences.  While
some argue that live coding must be performed live in front of an audience, an alternate
perspective suggests that time constraints are not necessary and that the practice can refer to any
improvisational music created by programming. The practice of live coding is still quite new.
For example, Nilson writes that professional violinists will typically have practiced for 10000 hours, where
as an email survey of live coders revealed most had dedicated about 100 hours. Of course, the
prerequisite knowledge for live coding may be higher; many live coders have extensive programming
experience. Also mentioned is the high level of abstraction offered by live coding creates a large
startup time for producing meaningful output from a blank slate.  Additionally, many exercises are
presented for improving one's live coding ability.

@@

Similar techniques~\cite{laroche1999new} enable new phase vocoder techniques that enable
exotic audio effects. The core of the techniques is a peak detection process followed by peak
shifting. Since these new techniques operate on peaks the standard phase vocoder algorithm can be
simplified to not require phase unwrapping, reducing computational complexity. Additionally, they
produce better quality output than the standard phase vocoder algorithm because of implicit
phase-locking.

@@

Bradford et al.\ have published a number of works detailing the sliding
DFT~\cite{bradford2005sliding} and associated sliding phase vocoder~\cite{bradford2007sliding}.
These new algorithms are efficient implementations for the STFT with a hop size of a single sample.
Although such a small hop size would be impossible to process in real-time using current hardware
and the STFT, the sliding DFT makes it possible using consumer GPU
hardware~\cite{bradford2011real}. The sliding phase vocoder additionally has a number of advantges
over traditional implementations. Quality is greatly improved but at a high computational cost. Each
analysis bin can hold any frequency, essentially making the spectrum more of a generic oscillator
bank than a traditional spectrum. This simplifies effects such as pitch shifting to a simple vector
multiplication in the frequency domain. Additionally, since the time resolution is so fine, effects
that require time modulation such as FM or AM can now be performed in the frequency domain.

@@

While the Fourier transform is by far the most commonly used technique for spectral manipulation in
digital audio effects, there exist numerous other transformations which may be useful as well to
achieve certain audio effects. One such technique is the Mellin transform, described in a audio
effects context by De Sena and Rocchesso~\cite{de2004fast}. The Mellin transform has a scale
invariance property that is analogous to the shift invariance property of the Fourier transform. It
can be computed quickly by applying two extra steps before doing a Fourier transform: exponential
time warping and exponential multiplication. A few simple audio effects are presented using the
Mellin transform, time stretching, filtering and random phase deviation (\textit{whisperization}).

@@

Collins et al.~\cite{collins2003live} present an introduction to the field of live coding where
code to generate sound is written live during a performance, often on a laptop computer. Although
the use graphical programming environments such as Reaktor, Pure Data, or MAX/MSP in live
performances is in some sense live coding, the paper focuses on the use of text based programming
environments. Many of the pros and cons of live coding are presented. Although live coding adds
some very interesting new possibilities to live music performance, it is not without it's
challenges. Case studies of two live coding environments are included: Slub and SuperCollider with
the \textit{just in time} library.

@@

This review reveals many of the possible options for the DSL used within the project. Although we
decided not to design a DSL for our project, analyzing the design and implementation techniques
presented here will be helpful for future work in this area.

@@

Since the ixi lang is similar to ScPy in many ways, many of the design choices made in ixi lang are
also appropriate in ScPy. Lessons learned from the user feedback for ixi lang could
potentially be used to improve usability of ScPy.

@@

def fn(x):
    x = pv.forward(x)
    x = pv.to_bin_offset(x)
    x.imag = -x.imag # imaginary part holds frequency
    x = pv.from_bin_offset(x)
    x = pv.backward(x)
    return x
    \end{minted}
    \caption{An effect which inverts the frequency of each analysis bin across the center of
    the bin, creating an unusual detuning effect.}
    \label{lst:binflipper}
\end{listing}

@@

Control UGens only provide a single floating point value per cycle so it would seem natural to
convert them to a Python \texttt{float}, but in ScPy they are instead transferred to Python in a
one element \texttt{ndarray}. This is because Python provides no numerical types with interior
mutability. In practice, NumPy accepts a one element \texttt{ndarray} almost anywhere a
\texttt{float} is accepted, so this detail is not an annoyance.

@@

As shown in \autoref{lst:shepard} and \autoref{lst:binflipper}, ScPy includes a phase vocoder
implemented in Python for performing frequency transformations. Interestingly, although all of
SuperCollider's FFT operations are referred to as phase vocoder operations, there is actually no
implementation of a phase vocoder to be found in the SuperCollider code base at the time of
writing. It is likely that a phase vocoder was once intended to be implemented, but the 32
so-called phase vocoder effects are simply transformations to the Cartesian or polar
representations of the spectrum.

@@

At the time of writing, ScPy only supports handling audio rate data via SuperCollider's
\texttt{Buffer} objects. In order to make some use cases more ergonomic ScPy could certainly be
extended to support raw audio rate input. For example, this would make time domain audio data even
simpler to work with than frequency domain data is currently. It would also enable SuperCollider's
FFT UGen to be easily re-implemented in Python for greater flexibility.

@@

Throughout~\autoref{sec:lit}, articles are discussed which contain processing techniques that could
be used in ScPy. Although implementations of many of these techniques are likely to become a part
of ScPy in the future, there is no obstacle to the reader implementing them as Python libraries
usable by ScPy or even as contributions to the project.

@@

While the number of future extensions to ScPy is numerous, the basic implementation that currently
exists is more than enough to yield countless exotic and novel audio effects. In the limited
experiments that we did using the software, it was found to be very easy to produce many advanced
sounds, thanks to the excellent documentation and implementation of the Python language and NumPy
library. Although it was possible to make these effects without our technology, ScPy makes it
possible to experiment with them at an unprecedented rate and listen to their output in real-time.

>>>>

Nilson~\cite{nilson2007live} presents a review of the challenges of live coding practice. Live
coding presents an interesting contrast to contemporary music with many clear differences.  While
some argue that live coding must be performed live in front of an audience, an alternate
perspective suggests that time constraints are not necessary and that the practice can refer to any
improvisational music created by programming. The practice of live coding is still quite new.
For example, Nilson writes that professional violinists will typically have practiced for 10000 hours, where
as an email survey of live coders revealed most had dedicated about 100 hours. Of course, the
prerequisite knowledge for live coding may be higher; many live coders have extensive programming
experience. Also mentioned is that the high level of abstraction offered by live coding creates a large
startup time for producing meaningful output from a blank slate. The paper presents many exercises
for improving one's live coding ability.

@@

Similar techniques also by Laroche and Dolson~\cite{laroche1999new} enable new phase vocoder
techniques that can be used for exotic audio effects. The core of the techniques is a peak
detection process followed by peak shifting. Since these new techniques operate on peaks the
standard phase vocoder algorithm can be simplified to not require phase unwrapping, reducing
computational complexity. Additionally, they produce better quality output than the standard phase
vocoder algorithm because of implicit phase-locking.

@@

Bradford et al.\ have published a number of works detailing the sliding
DFT~\cite{bradford2005sliding} and associated sliding phase vocoder~\cite{bradford2007sliding}.
These new algorithms are efficient implementations for the STFT with a hop size of a single sample.
Although such a small hop size would be impossible to process in real-time using current hardware
and the STFT, the sliding DFT makes it possible using consumer GPU
hardware~\cite{bradford2011real}. The sliding phase vocoder additionally has a number of advantges
over traditional implementations. Quality is greatly improved but at a high computational cost. Each
analysis bin can hold any frequency, essentially making the spectrum more of a generic oscillator
bank than a traditional spectrum. This simplifies effects such as pitch shifting to a simple vector
multiplication in the frequency domain. Additionally, since the time resolution is so fine, effects
that require high resolution modulation through time such as FM or AM can now be performed in the
frequency domain.

@@

While the Fourier transform is by far the most commonly used technique for spectral manipulation in
digital audio effects, there exist numerous other transformations which may be useful as well to
achieve certain audio effects. One such technique is the Mellin transform, described in a audio
effects context by De Sena and Rocchesso~\cite{de2004fast}. The Mellin transform has a scale
invariance property that is analogous to the shift invariance property of the Fourier transform. It
can be computed quickly by applying two extra steps before doing a Fourier transform: exponential
time warping and exponential multiplication. A few simple audio effects are presented using the
Mellin transform: time stretching, filtering and random phase deviation (\textit{whisperization}).

@@

Collins et al.~\cite{collins2003live} present an introduction to the field of live coding where
code to generate sound is written live during a performance, often on a laptop computer. Although
the use graphical programming environments such as Reaktor, Pure Data, or MAX/MSP in live
performances is in some sense live coding, the paper focuses on the use of text based programming
environments. Many of the pros and cons of live coding are presented. Although live coding adds
some very interesting new possibilities to live music performance, it is not without its
challenges. Case studies of two live coding environments are included: Slub and SuperCollider with
the \textit{just in time} library.

@@

This review presents many useful ideas for language design within ScPy. Although we decided not to
design a DSL for our project, analyzing the design and implementation techniques presented here
will be helpful for future work in this area.

@@

Since the ixi lang is similar to ScPy in many ways, many of the design choices made in ixi lang are
also appropriate in ScPy. Lessons learned from the user feedback for ixi lang could
potentially be used to improve the usability of ScPy.

@@

def fn(x):
    x = pv.forward(x)
    x = pv.to_bin_offset(x)
    x.imag *= -1 # imaginary part holds frequency
    x = pv.from_bin_offset(x)
    x = pv.backward(x)
    return x
    \end{minted}
    \caption{An effect which inverts the frequency of each analysis bin across the center of
    the bin, creating an unusual detuning effect.}
    \label{lst:binflipper}
\end{listing}

@@

Control UGens only provide a single floating point value per cycle so it would seem natural to
convert them to a Python \texttt{float}, but in ScPy they are instead transferred to Python in a
one element \texttt{ndarray}. This is because Python provides no numerical types with interior
mutability. In practice, NumPy accepts a one element \texttt{ndarray} almost anywhere a
\texttt{float} is accepted, so this detail is usually not an annoyance.

@@

As shown in \autoref{lst:shepard} and \autoref{lst:binflipper}, ScPy includes a phase vocoder
implemented in Python for performing frequency transformations. Interestingly, although all of
SuperCollider's FFT operations are referred to as phase vocoder operations in the code and
documentation, there is actually no implementation of a phase vocoder to be found in the
SuperCollider code base at the time of writing. It is likely that a phase vocoder was once intended
to be implemented, but the 32 so-called phase vocoder effects are simply transformations to the
Cartesian or polar representations of the spectrum.

@@

At the time of writing, ScPy only supports handling audio rate data via SuperCollider's
\texttt{Buffer} objects. In order to make some use cases more ergonomic ScPy could certainly be
extended to support raw audio rate input. This would make time domain audio data even
simpler to work with than frequency domain data is currently. It would also enable SuperCollider's
FFT UGen to be easily re-implemented in Python for greater flexibility.

@@

Throughout~\autoref{sec:lit}, articles are discussed which contain processing techniques that could
be used in ScPy. Although implementations of many of these techniques may to become a part of ScPy
in the future, there is no obstacle to the reader implementing them as external Python libraries
usable by ScPy.

@@

While the number of possible future extensions to ScPy is numerous, the basic implementation that
currently exists is more than enough to yield countless exotic and novel audio effects. In the
limited experiments that we did using the software, it was found to be very easy to produce many
advanced sounds, thanks to the excellent documentation and implementation of the Python language
and NumPy library. Although it was possible to make these effects without our technology, ScPy
makes it possible to experiment with them at an unprecedented rate and listen to their output in
real-time.
